Edge effects in Alberta

### 2021-07-27 Manuscript/manuscript.Rmd
Worked on manuscript and data filtering

- SNL in and around crops is important for both agricultural production and conservation
  - They are habitat for mobile organisms, and can therefore act as sources of ecosystem services such as pollination or pest control
  - They also can create microclimate effects that reduce extreme temperature, trap moisture, and reduce wind speed
  - Unfortunately, most of the research is concentrated in Europe, and they tend to be less-studied on other continents
  - In particular, North American agroecosystems have larger fields, and different varieties and agronomic practices, all of which could negate effects of SNL

<!-- - SNL has conflicting effects on crop yield because of ecosystem service provision as well as negative edge effects -->
- SNL may affect yields at intermediate distances, depending on the spatial scale at which ecosystem services operate
  - Edge effects cause low yields at the edge of crops because of sparse or late seedling emergence, poor microclimate, and competition with weeds
  - At the same time, the centre of large fields will not receive ecosystem services if they decay with distance from edges
  - For example, pollination services from central place foragers that nest in SNL but forage in crops drops rapidly with distance
  - Therefore, yield may be maximized at intermediate distances, where the ecosystem services cancel out negative edge effects 
  - This suggests a ``goldilocks" field size, where negative edge effects are canceled out by ecosystem services

@@

- Studies of SNL effects on crop yield also suffer from limited scope (e.g. few crop types) and small sample sizes, limiting inference and reducing generality (but see wheat study)
  - For this reason, large-scale precision yield data holds enormous promise for agronomy, as it allows 
  - However, its use is limited for several reasons: 1) Lack of standardized formats between equipment types, 2) Sensor calibration required for field-level accuracy, and 3) unfamiliarity with spatial statistics
  - Ecosystem services can influence both the mean and variability of yield in agroecosystems
  - Typically only averages (means) are considered, but higher stability (lower variance) in yield can also be valuable
  - Size examples: wheat study from the UK
  - There are few studies of yield variability (only those with large datasets), but precision yield data opens up the possibility of modeling within-field variability, as well as average yield

@@

- Precision yield data were collected directly from farmers across Alberta
  - Farmers were solicited for yield data through local agronomists, and we received `r nrow(datSource)` field-years of data from `r length(unique(datSource$grower))` growers across a total of `r length(unique(datSource$year))` years (`r min(unique(datSource$year))`-`r max(unique(datSource$year))`)
  - We converted data to a standard csv format using Ag Leader SMS
  - `r round(100*sum(grepl('(Wheat|Canola)',datSource$crop))/nrow(datSource))`% of the crop types where either wheat (_Triticum aestivum_) or canola (_Brassica napus_), two of the most common crops in rotation in Alberta
  - The remaining crop types were poorly replicated in our sample, so we constrained our analysis to only field-years containing wheat (`r sum(datSourceFinal$crop=='Wheat')`) or canola (`r sum(datSourceFinal$crop=='Canola')`)
  - Individual fields contained between `r min(yearsPerField)` to `r max(yearsPerField)` years of data (mean: `r round(mean(yearsPerField),1)`)
  - Containing a total of `r round(sum(datSource$npoints)/1e6,1)` million data points

- Yield data is collected in rectangles of the same length as the data interval (distance = combine ground speed $\times$ interval, typically 1 second) and the same width as the combine header (5-7 m)
  - We extracted the size of each polygon (m$^2$), dry yield (tonnes), and the spatial location, and the sequence of collection (1 - end of harvest)
  - Because of the large number of yield rectangles per field (30-800 thousand), we used the centroid of each polygon as its location, treating areal data as point data
  - Seeding and application rates were constant across fields, so we did not consider inputs in our analysis
  - We used dry yield (tonnes of seed/hectare after accounting for crop moisture) as our measure of crop yield
  - Due to the large number of data at each field, we sub-sampled to 50,000 data points per field to reduce computation time

@@

- At each field, we fit an additive model of the effect of boundary distance on crop yield while accounting for within-field spatial variation and temporal variation in the combine yield monitor
  - Crop yield can vary within a field due to soil conditions, moisture, seeding rates, herbicide application, and previous agricultural practices 
    <!-- - such as strip farming -->
  - Ground speed is extremely important to yield monitor accuracy [@arslan2002], with low ground speed registering higher yields 
  - While sensor calibration can reduce combine-level bias (such as a combine recording consistently higher/lower yields across fields), this does not address sensor drift that occurs over time  within fields
  - This may be caused by sensors accumulating debris during harvest (pers. comm. Trent Clark), leading to changes in accuracy and bias over time

@@

- In addition to modeling edge effects (our variable of interest), this model also accounts for a) differences in combine speed, b) within-field spatial variation not related to edges, and d) shifts in combine accuracy during harvest
  - Spatial or temporal variation is typically modelled using a Gaussian Process Model (Kriging) or approximations such as Stochastic Partial Differential Equations (e.g. INLA), but this was computationally infeasible with 50,000 data points per field
  - Penalized splines offer a compromise, as they account for nonlinear ``wiggly" relationships in the same way as Gaussian processes but with reduced dimensionality and computation time
  - The number of basis dimensions was checked with the _gam.check_ function from _mgcv_
  - The relationship between polygon size (i.e. ground speed) and yield was modeled with a log-linear relationship with a single slope term, as this closely matched smoothed versions
  - All models were fit in _R_ using the _mgcv_ library [version `r packageVersion('mgcv')`, @wood2017], and figures were created with _ggplot2_ and _ggpubr_ [versions `r packageVersion('ggplot2')`, @wickham2016; and `r packageVersion('ggpubr')`, @kassambara2020].

- To consider results from all field-level models, we fit models independent of each other, and "overall" smoothers were taken as averages of the field-level smoothers
  - However, this does not account for uncertainty in the field-level smoothers, so we used an approach similar to bootstrapping of hierarchical mixed effects models
  1. Extract single posterior sample (_rnorm_ in R) of smoother parameters from each field-level model using coefficient estimates and standard error
  2. Use posterior sample to create new simulated smoother from each field
  3. Fit new model of simulated smoothers from all fields, and save this "meta-smoother"
  4. Repeat 1000 times, and calculate coverage intervals (5-95% percentiles) on saved meta-smoothers
  - This gives coverage intervals (CIs) for the "average" smoother while accounting for field-level variability

>>>>

- SNL in and around crops is important for both agricultural production and conservation
  - They are habitat for mobile organisms, and can therefore act as sources of ecosystem services such as pollination or pest control (albrecht2020)
  - They also can create microclimate effects that reduce extreme temperature, trap moisture, and reduce wind speed (kort1988)
  - Unfortunately, most of the research is concentrated in Europe, and they tend to be less-studied on other continents
  - In particular, North American agroecosystems have larger fields, different crop varieties, and agronomic practices, all of which could negate or confound potential effects of SNL

<!-- - SNL has conflicting effects on crop yield because of ecosystem service provision as well as negative edge effects -->
- SNL may affect yields at intermediate distances, depending on the spatial scale at which ecosystem services operate
  - Edge effects cause low yields at the edge of crops because of sparse or late seedling emergence, poor microclimate, and competition with weeds
  - At the same time, the centre of large fields will not receive ecosystem services if they decay with distance from edges (kowalchuk1995)
  - For example, pollination services from central place foragers that nest in SNL but forage in crops drops rapidly with distance
  - Therefore, yield may be maximized at intermediate distances, where the ecosystem services cancel out negative edge effects (e.g. vanVooren2017)
  - This suggests a ``goldilocks" field size, where negative edge effects are canceled out by ecosystem services

@@

- Studies of SNL effects on crop yield also suffer from limited scope (e.g. few crop types) and small sample sizes, limiting inference and reducing generality (but see wheat study)
  - Meta-analyses have been the standard solution to this problem, but precision yield data holds enormous promise
  - However, its use is limited for several reasons: 1) Lack of standardized formats between equipment types, 2) Sensor calibration required for field-level accuracy, and 3) unfamiliarity with spatial statistics
  - Ecosystem services can influence both the mean and variability of yield in agroecosystems (redhead2020)
  - Typically only averages (means) are considered, but higher stability (lower variance) in yield can also be valuable
  - There are few studies of yield variability (only those with large datasets), but precision yield data opens up the possibility of modeling within-field variability, as well as average yield

@@

- Precision yield data were collected directly from farmers across Alberta
  - Farmers were solicited for yield data through local agronomists, and we received `r nrow(datSource)` field-years of data from `r length(unique(datSource$grower))` growers across a total of `r length(unique(datSource$year))` years (`r min(unique(datSource$year))`-`r max(unique(datSource$year))`)
  - We converted raw data to a standard csv format using Ag Leader SMS
  - `r round(100*sum(grepl('(Wheat|Canola)',datSource$crop))/nrow(datSource))`% of the crop types where either wheat (_Triticum aestivum_) or canola (_Brassica napus_), two of the most common crops in rotation in Alberta
  - The remaining crop types had low replication, so we constrained our analysis to only field-years containing wheat (`r sum(datSourceFinal$crop=='Wheat')`) or canola (`r sum(datSourceFinal$crop=='Canola')`)
  - Individual fields contained between `r min(yearsPerField)` to `r max(yearsPerField)` years of data (mean: `r round(mean(yearsPerField),1)`)
  - Containing a total of `r round(sum(datSource$npoints)/1e6,1)` million data points

- Yield data is collected in rectangles of the same length as the data interval (distance = combine ground speed $\times$ interval, typically 1 second) and the same width as the combine header (5-7 m)
  - We extracted the size of each polygon (m$^2$), dry yield (tonnes), and the spatial location, and the sequence of collection (1 - end of harvest)
  - Because of the large number of yield rectangles per field (30-800 thousand), we used the centroid of each polygon as its location, treating areal data as point data
  - Seeding and application rates were constant across fields, so we did not consider inputs in our analysis
  - We used dry yield (tonnes of seed/hectare minus recorded crop moisture) as our measure of crop yield
  - Precision yield data can be highly variable and is prone to extreme outliers (both positive and negative), especially when combine ground speeds are low and at the beginning of rows
  - Therefore, we filtered data that lay outside the 95th percentiles of yield within each field, values less than or equal to zero, and values outside of the 95th percentiles of ground speed (high: 
  - Due to the large number of data at each field, we sub-sampled to 50,000 data points per field to reduce computation time

@@

- Yield data from each field contains both random and systematic errors that must be modeled in order to reveal underlying changes in yield
  - We fit an additive model of the effect of boundary distance on crop yield while accounting for within-field spatial variation and temporal variation in the data
  - Crop yield can vary within a field due to soil conditions, moisture, seeding rates, herbicide application, and previous agricultural practices 
    <!-- - such as strip farming -->
  - Ground speed is extremely important to yield monitor accuracy [@arslan2002], with low ground speed registering higher yields 
  - Sensor calibration can reduce combine-level bias (such as a combine recording consistently higher/lower yields across fields), this does not address sensor drift that occurs over time
  - This may be caused by sensors accumulating debris during harvest (pers. comm. Trent Clark), leading to changes in accuracy and bias over time

@@

- In addition to modeling edge effects (our variable of interest), this model also accounts for a) differences in combine speed, b) within-field spatial variation not related to edges, and d) shifts in combine accuracy during harvest
  - Spatial or temporal variation is typically modelled using a Gaussian Process Model (e.g. Kriging, SPDE Approximations) but this was computationally unfeasible with 50,000 data points per field
  - Penalized splines offer a compromise, as they account for nonlinear ``wiggly" relationships in the same way as Gaussian processes but using a reduced number of dimensions (as well as computation time)
  - The number of basis dimensions was checked with the _gam.check_ function from _mgcv_
  - The relationship between polygon size (i.e. ground speed) and yield was modeled with a log-linear relationship with a single slope term, as this closely matched smoothed versions
  - All models were fit in _R_ using the _mgcv_ library [version `r packageVersion('mgcv')`, @wood2017], and figures were created with _ggplot2_ and _ggpubr_ [versions `r packageVersion('ggplot2')`, @wickham2016; and `r packageVersion('ggpubr')`, @kassambara2020].

- To consider results from all field-level models together, we fit models independent of each other and fit an "overall" smoother as averages of the field-level smoothers
  - However, this does not account for uncertainty in the field-level smoothers, so we used an approach similar to bootstrapping of hierarchical mixed effects models
  1. Extract single posterior sample (_rnorm_ in R) of smoother parameters from each field-level model using coefficient estimates and standard error
  2. Use posterior sample to simulated a new smoother from each field
  3. Fit new model of simulated smoothers from all fields, and save this "meta-smoother"
  4. Repeat 1000 times, and calculate coverage intervals (5-95% percentiles) on saved meta-smoothers
  - This gives coverage intervals (CIs) for the "average" smoother while accounting for field-level variability
